---
title: "exercise-12"
output: html_document
date: "2024-04-10"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Preliminaries
- Using the {tidyverse} read_tsv() function, load the “Mammal_lifehistories_v2.txt” dataset from this [URL](https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/Mammal_lifehistories_v2.txt) as a “tibble” named d. As discussed in class, this is dataset that compiles life history and other variables for over 1400 species of placental mammals from 17 different Orders.
- Do a bit of exploratory data analysis with this dataset, e.g., using the {skimr} package. Which of the variables are categorical and which are numeric?

    - Categorical variables: order, family, Genus, and species
    - Numeric variables: mass(g), gestation(mo), newborn(g), weaning(mo), wean mass(g), AFR(mo) max.life(mo), litter size, litters/year, refs

```{r}
# Load libraries 
library(skimr)
library(tidyverse)
library(naniar)

# Grab file
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/Mammal_lifehistories_v2.txt"
d <- read_tsv(f, col_names = TRUE)
skim (d)
head(d)

```
#### Step 1 
- Replace all values of -999 (the authors’ code for missing data) with NA.

```{r}
# use replace_with_na_all() function from the {naniar} to replace -999 with NA.
d <- replace_with_na_all(data = d, condition = ~.x == -999.00)

```

#### Step 2 
- Drop the variables **litter size** and **refs**.

```{r}
d <- d %>%
  dplyr::select(-c(`litter size`, refs))

# double check 
glimpse(d) # litter size and refs are dropped 
```

#### Step 3
- Log transform all of the other numeric variables

```{r}
# The mutate(across(where(is.numeric))) looks for variables that are numeric and log transforms. NOTE: Variable names DO NOT CHANGE. I am keeping them the same. Option to mutate(rename())
d <- d %>% 
  mutate(across(where(is.numeric), ~log(.)))

glimpse(d)
```


#### Step 4 
- Regress the (now log transformed) age [gestation(mo), weaning(mo), AFR(mo) (age at first reproduction), and max. life(mo) (maximum lifespan)] and mass [newborn(g) and wean mass(g)] variables on (now log transformed) overall body mass(g) and add the residuals to the dataframe as new variables (relGest, relWean, relAFR, relLife, relNewbornMass, and relWeaningMass).
```{r}
# Regress the selected variables on log transformed body mass, called mass(g)
# Must include na.action = na.include in order to include NA to make sure the matrix fits 
m1 <- lm(`mass(g)` ~ `gestation(mo)`, data = d, na.action = na.exclude)
m2 <- lm(`mass(g)` ~ `weaning(mo)`, data = d, na.action = na.exclude)
m3 <- lm(`mass(g)` ~ `AFR(mo)`, data = d, na.action = na.exclude)
m4 <- lm(`mass(g)` ~ `max. life(mo)`, data = d, na.action = na.exclude)
m5 <- lm(`mass(g)` ~ `newborn(g)`, data = d, na.action = na.exclude)
m6 <- lm(`mass(g)` ~ `wean mass(g)`, data = d,  na.action = na.exclude)


# Need to call resid(). Doing m1$residuals did not include NA...
d <- d %>%
  mutate(
    relGest = resid(m1),
    relWean = resid(m2), 
    relAFR = resid(m3),
    relLife = resid(m4),
    relNewbornMass = resid(m5), 
    relWeaningMass = resid(m6)
  )

```

#### Step 5
- Plot residuals of max lifespan (relLife) in relation to Order. Which mammalian orders have the highest residual lifespan?

    - Based on maximum values, the mammalian orders with the highest residual lifespan: (1) Sirenia, (2) Perissodactyla, and (3) Rodentia
    - Based on median values, the mammalian orders with the highest residual lifespan: (1) Cetacea, (2) Artiodactyla, and (3) Proboscidea
    
```{r}
# Plot: Residuals of max lifespan (relLife) in relation to Order
p_relLife <- ggplot(data = d %>% drop_na(relLife), aes(x = order, y = relLife)) + 
  geom_boxplot() + 
  geom_jitter(alpha = 0.05) + # makes the dots more transparent 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Order") + ylab("Residual Log(Max Lifespan)")
p_relLife

# To double check: max lifespan::order; use R's 5 number summary function fivenum() to definitively tell which orders have the highest max
relLife_5NumSum <- d %>%
  group_by(order) %>%
  summarise(n = n(),
            min = fivenum(relLife)[1],
            Q1 = fivenum(relLife)[2],
            median = fivenum(relLife)[3],
            Q3 = fivenum(relLife)[4],
            max = fivenum(relLife)[5])
relLife_5NumSum
```

- Plot residuals of newborn mass (relNewbornMass) in relation to Order. Which mammalian orders have the have highest residual newborn mass?

    - Based on maximum values, the mammalian orders with the highest residual newborn mass: (1) Carnivora (2) Xenarthra (3) Artiodactyla
    - Based on median values, the mammalian orders with the highest residual newborn mass: (1) Carnivora (2) Tubulidentata (3) Proboscidea
    
```{r}    
# Plot: Residuals of newborn mass (relNewbornMass) in relation to Order
p_relNewbornMass <- ggplot(data = d %>% drop_na(relNewbornMass), aes(x = order, y = relNewbornMass)) + 
  geom_boxplot() + 
  geom_jitter(alpha = 0.05) + # makes the dots more transparent 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Order") + ylab("Residual Log(Newborn Mass (g)")
p_relNewbornMass

# It is difficult to tell what some of the maxes were on the graph, so using R's 5 number summary function fivenum() to definitively tell which orders have the highest max
relNewbornMass_5NumSum <- d %>%
  group_by(order) %>%
  summarise(n = n(),
            min = fivenum(relNewbornMass)[1],
            Q1 = fivenum(relNewbornMass)[2],
            median = fivenum(relNewbornMass)[3],
            Q3 = fivenum(relNewbornMass)[4],
            max = fivenum(relNewbornMass)[5])
relNewbornMass_5NumSum
```
    
- Plot residuals of weaning mass (relWeaningMass) in relation to Order. Which mammalian orders have the have highest relative weaning mass?

    - Based on maximum values, mammalian orders with the highest relative weaning mass: (1) Carnivora, (2) Rodentia, (3) Lagomorpha
    - Based on median values, the mammalian orders with the highest relative weaning mass: (1) Xenarthra, (2) Tubulidentata, (3) Hyracoidea
    
```{r}    

# Plot: Residuals of weaning mass (relWeaningMass) in relation to Order
p_relWeaningMass <- ggplot(data = d %>% drop_na(relWeaningMass), aes(x = order, y = relWeaningMass)) + 
  geom_boxplot() + 
  geom_jitter(alpha = 0.05) + # makes the dots more transparent 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Order") + ylab("Residual Log(Weaning Mass (g))")
p_relWeaningMass

# To double check plots 
relWeaningMass_5NumSum <- d %>%
  group_by(order) %>%
  summarise(n = n(),
            min = fivenum(relWeaningMass)[1],
            Q1 = fivenum(relWeaningMass)[2],
            median = fivenum(relWeaningMass)[3],
            Q3 = fivenum(relWeaningMass)[4],
            max = fivenum(relWeaningMass)[5])
relWeaningMass_5NumSum
```

#### Step 6
- Run models and a model selection process to evaluate what (now log transformed) variables best predict each of the two response variables, max. life(mo) and AFR(mo), from the set of the following predictors: gestation(mo), newborn(g), weaning(mo), wean mass(g), litters/year, and overall body mass(g).
- For each of the two response variables, indicate what is the best model overall based on AICc and how many models have a delta AICc of 4 or less.

    - For `max. life(mo)` m4 is the best model which includes log transformed variables of `gestation(mo)`, `litters/year`, `mass(g)`,`weaning(mo)`. There are 5 models that have a delta AICc of 4 or less.
    - For `AFR(m0)`, the forward selection model has the log transformed variables of `litters/year`, `gestation(mo)`, `weaning(mo)`, and `wean mass(g)` as adding explanatory power. Reverse selection and AIC models have the log transformed variables of `litters/year`, `gestation(mo)`, `weaning(mo)`, and `mass(g)` as adding explanatory power. There are 7 models that have a delta AICc of 4 or less. 
    
- What variables, if any, appear in all of this set of “top” models?

    - The variables that appear in all the sets are log transformed `gestation(mo)`, `litters/year`, and `weaning(mo)`. `mass(g)` appears in most models. 
    
- Calculate and plot the model-averaged coefficients and their CIs across this set of top models.

First: max life
```{r}
# First, drop NAs
# Drop NAs for the response variable max. life(mo)
d_maxLife <- d %>%
  drop_na(`max. life(mo)`,`gestation(mo)`, `newborn(g)`, `weaning(mo)`, `wean mass(g)`, `litters/year`, `mass(g)`)

# Forward Selection Regression Modeling for max. life(mo)
# max. life(mo) null model 
m0 <- lm(data = d_maxLife, `max. life(mo)` ~ 1)
add1(m0, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, test = "F")
# All values would add explanatory power, but gestation(mo) is associated with the highest new F statistic

m1 <- update(m0, formula = . ~ . + `gestation(mo)`)
summary(m1)

add1(m1, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, test = "F")
# All values would add explanatory power, but `litters/year` is associated with the highest new F statistic
m2 <- update(m1, formula = . ~ . + `litters/year`)
summary(m2)

add1(m2, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, test = "F")
# Newborn(g), wean mass(g), and mass(g) are all significant. Mass(g) has highest F value
m3 <- update(m2, formula = . ~ . + `mass(g)`)
summary(m3)

add1(m3, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, test = "F")
# Only weaning(mo) is significant!
m4 <- update(m3, formula = . ~ . + `weaning(mo)`)
summary(m4)

add1(m4, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, test = "F")
# No more terms are significant! There are not any more predictors that would improve the explanatory power of the model.
# m4 is our best model for max life (mo)

# ...Using AIC to select
library(MASS)

# Start with the full model 
m.full.MaxLife <- lm(data = d_maxLife, `max. life(mo)` ~ `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`)
s <- stepAIC(m.full.MaxLife, scope = .~., direction = "both")
summary(s)
# ... same as our m4 model: `gestation(mo)`, `weaning(mo)`, `litters/year`, and `mass(g)` add explanatory power to the model. 

# Check delta AICc of 4 or less
library(AICcmodavg)
aictab(list(m0, m1, m2, m3, m4), c("m0", "m1", "m2", "m3", "m4"))
```

Next: AFR 
```{r}
# Repeat for AFR (mo)
# Drop NAs for the response variable AFR(mo)
d_AFR <- d %>%
  drop_na(`AFR(mo)`,`mass(g)`, `gestation(mo)`, `newborn(g)`, `weaning(mo)`, `wean mass(g)`, `litters/year`, )

# Create AFR null linear model 
m0_AFR <- lm(data = d_AFR, `AFR(mo)` ~ 1)
add1(m0_AFR, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, test = "F")

# Create AFR null linear model
m0_AFR <- lm(data = d_AFR, `AFR(mo)`~ 1)

# Add all variables 
add1(m0_AFR, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)`+ `litters/year` + `mass(g)`, test = "F")

# All values would add explanatory power, but `litters/year` is associated with the highest new F statistic

m1_AFR <- update(m0_AFR, formula = . ~ . + `litters/year`)
summary(m1_AFR)

add1(m1_AFR, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)`+ `litters/year` + `mass(g)`, test = "F")

# All values would add explanatory power, but `gestation(mo)` is associated with the highest new F statistic

m2_AFR <- update(m1_AFR, formula = . ~ . + `gestation(mo)`)
summary(m2_AFR)

add1(m2_AFR, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)`+ `litters/year` + `mass(g)`, test = "F")

# `weaning(mo)`, `wean mass(g)`, and `mass(g)` would add explanatory power, but `wean mass(g)` is associated with the highest new F statistic

m3_AFR <- update(m2_AFR, formula = . ~ . + `wean mass(g)`)
summary(m3_AFR)

add1(m3_AFR, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)`+ `litters/year` + `mass(g)`, test = "F")

# Only `weaning (mo)` is associated with the highest new F statistic
m4_AFR <- update(m3_AFR, formula = . ~ . + `weaning(mo)`)
summary(m4_AFR)

add1(m4_AFR, scope = . ~ . + `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)`+ `litters/year` + `mass(g)`, test = "F")
#Remaining variables do not add any more explanatory power. 

# Same thing but backwards selection to double check 
full_AFR <- lm(data = d_AFR, `AFR(mo)` ~ `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`)
summary(full_AFR)

drop1(full_AFR, test = "F") 
# `wean mass(g)`  has non-sig p-value, lowest F, and lowest increase in RSS; does not add any explanatory power

m1_AFR.drop <- update(full_AFR, . ~ . - `wean mass(g)`)
summary(m1_AFR.drop)

drop1(m1_AFR.drop, test = "F") # `newborn(g)` has non-sig p-value, lowest F, and lowest increase in RSS; does not add any explanatory power

m2_AFR.drop <- update(m1_AFR.drop, . ~ . - `newborn(g)`)
summary(m2_AFR.drop) # all explanatory variables are significant, model 2 is best

aictab(list(m0_AFR, m1_AFR, m2_AFR, m3_AFR, m4_AFR), c("m0_AFR", "m1_AFR", "m2_AFR", "m3_AFR", "m4_AFR"))
detach(package:AICcmodavg)

# ...Using AIC to select
# Start with the full model 
m.full.AFR <- lm(data = d_AFR, `AFR(mo)` ~ `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, na.action = na.fail)
s <- stepAIC(m.full.AFR, scope = . ~ ., direction = "both")
summary(s)
# Forward selection and reverse selection yielded different results. Reverse selection had the same variables as AIC. `litters/year`, `gestation(mo)`, and  `weaning (mo)` are the same across and add explanatory power to the model. 
detach(package:MASS)
```

```{r}
# Using {MuMIn} package, we calculate the model-averaged coefficients and their CIs across this set of top models.
library(MuMIn)
# Max life (mo)
# Using full data set
m.full <- lm(data = d_maxLife, `max. life(mo)` ~ `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, na.action = "na.fail")

# dredge from {MuMIn} explores subsets of a given global model in an automated way.
mods <- dredge(m.full)

# number of models of delta AICc of 4 of less
(mods_delta <- subset(mods, delta <= 4)) 

# dredge() does not actually store the model results; use model.avg() function returns a “summary.averaging” object.
# average beta coefficients for models in the 95% confidence set, i.e., for cumulative AIC weights up to 95%:
(mods.avg <- summary(model.avg(mods, subset = delta <= 4, fit = TRUE)))

mods.avg$msTable

#use the confint() function for confidence intervals for the averaged average beta coefficients for models in the 95% CI set
confint(mods.avg)

# Using {MuMIn} package, we can also plot this
plot(mods.avg, full = TRUE, intercept = FALSE, main = "Model-averaged Coefficients for Maximum Life (months)")

# Age at first reproduction (AFR)
m.full <- lm(data = d_AFR, `AFR(mo)` ~ `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, na.action = "na.fail")

# dredge from {MuMIn}
mods <- dredge(m.full)

# number of models of delta AICc of 4 of less
(mods_delta <- subset(mods, delta <= 4)) 

# average coefficients for top models where delta.aicc <= 4
(mods.avg <- summary(model.avg(mods, subset = delta <= 4, fit = TRUE)))

mods.avg$msTable

#use the confint() function for confidence intervals for the averaged average beta coefficients for models in the 95% CI set
confint(mods.avg)

# Using {MuMIn} package, we can also plot this
plot(mods.avg, full = TRUE, intercept = FALSE, main = "Model-averaged Coefficients for \nAge at First Reproduction (months)")
```




